Topological Equivariant Artist Model for Visualization Library Architecture

### 2024-12-03 paper/main.tex
relayed work

Although the Steven's scales were conceptualized as having group structure, the ordinal scale has a monoidal structure because partial orders ($\geq, \leq$) are not invertable. This means \textit{equivariance} cannot be used to test for structure preservation. Instead \textit{homomorphsim} can be used because it imposes fewer constraints on the underlying mathematical structure of the data.

\subsubsection{homomorphism}
\begin{figure}[H]
  \includegraphics[width=1\columnwidth]{tex/homomorphism.png}
  \caption{Encoding data as bar height using an inverse transform is not homomorphic because the largest number is mapped to the smallest bar while the max function returns the largest bar.}
  \label{fig:related-work:homomorphism}
\end{figure}

@@

\subsection{Topology}
\label{sec:related-work:continuity}
\begin{figure}[H]
  \includegraphics[width=1\columnwidth]{k_different_types.png}
  \caption{
  \note{add first panel which is table + all K type, second is scatter, etc...}
%
  \label{fig:related-work:continuity:ktypes}}
\end{figure}
\note{lead with homomorphism}
Borrowing structures from algebraic topology allows us to describe structure on individual elements in a way that holds true for the entire dataset, meaning we can use the same tools to check if structre is preserved regardless of whether the data fits in memory, is distributed, or is streaming. Our framework generalizesis because, informally, a topology $\mathcal{T}$ is a partitioning of a space into its subspaces in a way where those subspaces can be composed back into the space; the choice of partitions is how we encode the connectivity of the data and visualizations that we are interested in.

@@

Visual algorithms assume the topology of their input data, as described in taxonomies of visualization algorithms Chi\cite{chiTaxonomyVisualizationTechniques2000} and by Troy and M\"{o}ller \cite{toryRethinkingVisualizationHighlevel2004}, but generally do not verify that input structure. For example, a \texttt{line} algorithm often does not have a way to query whether a list of (x,y) coordinates is the distinct rows, the time series, or the list of stations in \autoref{fig:related-work:continuity:ktypes}. While plotting the time series as a continuous line would be correct, it would be incorrect for a visualization to indicate that the distinct rows or stations are connected in a 1D continuous manner because it introduces ambiguity over which part of the line maps back to the data. A map that by definition has continous maps between two spaces, such as data and graphics, is called a homeomorphism\cite{riehlCategoryTheoryContext}.
\note{add functional def of bijective and continous
}
\begin{definition}
  A function $f$ is a $homeomorphism$ if it is bijective, continous, and has a continous inverse function $f^{-1}$.
\end{definition}

To encode topology and field structure, we extend Butler's work on using a mathematical structure called fiber bundles as an abstract data representation in visualization \cite{butlerVectorBundleClassesForm1992, butlerVisualizationModelBased1989}. We sketch out fiber bundles in \autoref{sec:atct:fiber-bundles}, but Butler provides a thorough introduction to bundles for visualization practitioners.

\note{this got moved}
In \autoref{fig:related-work:continuity:ktypes}, the topology is a continuous 3 dimensional surface encapsulating time and position, while the fields are date, location, station, pressure, and temperature. Our paper provides a language to express how the scatter plot, line plot, and heatmap are preserving different structural properties of the dataset, thereby generalizing Bertin's codification of how the topology of observations must match the class of representation (i.e. point , line, area) and how the variation possible in a field match the variation in the corresponding visual variable \cite{bertinSemiologyGraphicsDiagrams2011} and Mackinlay's work on how structure preservation includes preserving the mathimatical relations between elements of the same and different fields \cite{mackinlayAutomatingDesignGraphical1986}.

@@

The coupling of values with their allowed transformations is roughly equivalent to the data and methods of and on a software class. Category theory refines the mapping between math and software and its use can lead to more consistent software interfaces in visualization software libraries \cite{wielsManagementEvolvingSpecifications1998,yorgeyMonoidsThemeVariations2012}. A brief visualization oriented introduction to category theory is in Vickers et al \cite{vickersUnderstandingVisualizationFormal2013}, but they are applying category theory to semantic concerns about visualization design rather than library architecture.

>>>>

Although the Steven's scales were conceptualized as having group structure, the ordinal scale has a monoidal structure because partial orders ($\geq, \leq$) are not invertable. This means \textit{equivariance} cannot be used to test for structure preservation. Instead \textit{homomorphsim} can be used because it imposes fewer constraints on the underlying mathematical structure of the data.
\begin{figure}[H]
  \includegraphics[width=1\columnwidth]{tex/homomorphism.png}
  \caption{Encoding data as bar height using an inverse transform is not homomorphic because the largest number is mapped to the smallest bar while the max function returns the largest bar.}
  \label{fig:related-work:homomorphism}
\end{figure}

@@

\subsection{Topology}
\label{sec:related-work:continuity}
Visual algorithms assume the topology of their input data, as described in taxonomies of visualization algorithms Chi\cite{chiTaxonomyVisualizationTechniques2000} and by Troy and M\"{o}ller \cite{toryRethinkingVisualizationHighlevel2004}, but generally do not verify that input structure.

\begin{figure}[]
  \includegraphics[width=1\columnwidth]{k_different_types.png}
  \caption{This weather station data has multiple embedded continuities - points at each time and position, timeseries at each position, and maps at each time. The corresponding visualizations - bar chart, timeseries, and map - each preserve the continuity of the subset of the data they visualize such that there is continuous map from the visualization (continuity) to the data (continuity).}
%
  \label{fig:related-work:continuity:ktypes}
\end{figure}

For example, a \texttt{line} algorithm often does not have a way to query whether a list of (x,y) coordinates is the distinct rows, the time series, or the list of stations in \autoref{fig:related-work:continuity:ktypes}. While plotting the time series as a continuous line would be correct, it would be incorrect for a visualization to indicate that the distinct rows or stations are connected in a 1D continuous manner because it introduces ambiguity over which part of the line maps back to the data. A map that by definition has continous maps between two spaces, such as data and graphics, is called a \textit{homeomorphism}\cite{riehlCategoryTheoryContext}:

\begin{definition}
  A function $f$ is a $homeomorphism$ if it is bijective, continous, and has a continous inverse function $f^{-1}$.
\end{definition}

Therefore the the bar plot, line plot, and heatmap in \autoref{fig:related-work:continuity:ktypes} have a homemeomorphic relationship to the 0D points, 1D timeseries, and 2D map continuities embedded in the continuous 3 dimensional surface encapsulating time and position. Using homeomorphism to test whether continuity is preserved generalizes Bertin's codification of how the topology of observations must match the class of representation (i.e. point , line, area) \cite{bertinSemiologyGraphicsDiagrams2011} and Wilkinson's assertion that connectivity be preserved \cite{wilkinsonGrammarGraphics2005}.

Describing connectivity using the language of topology allows for describing individual elements in a way that holds true for the entire dataset. This means that the framework described in this paper can be used regardless of whether the data fits in memory, is distributed, or is streaming, because, informally, a topology $\mathcal{T}$ is a partitioning of a space into its subspaces in a way where those subspaces can be composed back into the total space:

@@

This means that the data can be partitioned arbitrarily and still retain it's continuity; for example the topology of the data in \autoref{fig:fig:related-work:continuity:ktypes} could be encoded as the set of individual points (one at each time and station) or set of timeseries (one per station) or set of maps (one per time). In this topological model, semantic indexing as described by Munzner's key-value model of data structure act as different ways of building the partitions. It also makes clearer when different labeling schemes refer to the same point, for example how 0-360 and 180E-180W are two ways of labeling longitude. To encode topology and field structure in a way that is both uniform and generalizable, we extend Butler's work on using a mathematical structure called fiber bundles as an abstract data representation in visualization \cite{butlerVectorBundleClassesForm1992, butlerVisualizationModelBased1989}. We sketch out fiber bundles in \autoref{sec:atct:fiber-bundles}, but Butler provides a thorough introduction to bundles for visualization practitioners.

@@

We propose that the cohesion of domain specific library APIs is obtainable using the uniform data model described in \autoref{sec:atct:fiber-bundles} while the expressivity of building block libraries can be preserved by defining explicit structure preserving constraints on the library components, as described in \autoref{sec:artist}. Because category theory constructions map cleanly to objects and functions, using category theory to express the structure and constraints can lead to more consistent software interfaces in visualization software libraries \cite{wielsManagementEvolvingSpecifications1998,yorgeyMonoidsThemeVariations2012}. A brief visualization oriented introduction to category theory is in Vickers et al \cite{vickersUnderstandingVisualizationFormal2013}, but they are applying category theory to semantic concerns about visualization design rather than library architecture.
